Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-08 paper/main.tex
Updates
Udpates

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative method that guides
researchers in manually constructing or improving the fit of an existing DAG.
Our approach combines implied Conditional Independence (CI) testing with a
measure of conditional association between variables to rank violations
of implied CIs in a given DAG. This ranking helps prioritize modifications to
the DAG to improve its fit to the data while utilizing expert knowledge to
decide the orientation of new edges. Empirical results show that this guided
manual model construction approach achieves performance comparable to automated
algorithms if we are able to correctly decide the orientation of new edges in
one out of three cases. Additionally, we demonstrate that, in the absence of an
expert, a Large Language Model (LLM) can be potentially used to determine edge
orientations. We provide the implementation of our approach in both a web tool
at: <redacted for review> and a Python package <redacted for review>.

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\emph{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\emph{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative method that guides
researchers in manually constructing or improving the fit of an existing DAG
while incorporating their expert knowledge. Compared to other expert knowledge
specification methods, this method explicitly asks for the causal ordering
between two specific variables in each step. Additionally, we provide a way to
rank the potential new edge addition that improves the model fit the most in
each step. Empirical results show that this guided manual model construction
approach achieves performance comparable to automated algorithms if we are able
to correctly decide the orientation of new edges in one out of three cases.
Additionally, we demonstrate that, in the absence of an expert, a Large
Language Model (LLM) can be potentially used to determine edge orientations. We
provide the implementation of our approach in both a web tool at: <redacted for
review> and a Python package <redacted for review>.

% Our approach combines implied Conditional Independence (CI) testing with a
% measure of conditional association between variables to rank violations
% of implied CIs in a given DAG. This ranking helps prioritize modifications to
% the DAG to improve its fit to the data while utilizing expert knowledge to
% decide the orientation of new edges.

@@

\subsection{Ranking Potential New Edges}

\subsection{Comparison to Automated Algorithms}
\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\emph{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\emph{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}
